---
title: "AIC/BIC tests"
bibliography: references.bib
format:
  html:
    code-copy: true
editor_options: 
  chunk_output_type: console
---

## Background

Akaike’s Information Criterion (AIC) and the Bayesian Information Criterion (BIC) provide a useful statistical test of the relative fit of alternative parametric models, and they are usually available as outputs from statistical software. Further details on these are available from [@Collett2003]. Measures such as the negative 2 log likelihood are only suitable for comparing nested models, whereby one model is nested within another (for example, one model adds an additional covariate compared to another model). Different parametric models which use different probability distributions cannot be nested within one another. Thus the negative 2 log likelihood test is not suitable for assessing the fit of alternative parametric models. The AIC and BIC allow a comparison of models that do not have to be nested, including a term which penalises the use of unnecessary covariates (these are penalised more highly by the BIC). Generally, it is not necessary to include covariates in survival modelling in the context of an RCT as it would be expected that any important covariates would be balanced through the process of randomisation. However, some parametric models have more parameters than others, and the AIC and BIC take account of these – for example an exponential model only has one parameter and so in comparative terms two parameter models such as the Weibull or Gompertz models are penalised. The AIC and BIC statistics therefore weigh up the improved fit of models with the potentially inefficient use of additional parameters, with the use of additional parameters penalised more highly by the BIC relative to the AIC.


## R example

Cox proportional hazard model

```{r}
coxph()

AIC()
extractAIC()

BIC()
```

In a Bayesian model

```{r}
library(survHE)

survHE:::compute_ICs_stan()
```



