---
title: "Proportional Hazard Modelling"
bibliography: references.bib
format:
  html:
    code-copy: true
editor_options: 
  chunk_output_type: console
---

# Introduction

The purpose of the CPH model is to evaluate simultaneously the effect of several factors on survival. It allows us to examine how specified factors influence the rate of a particular event happening at a particular point in time. This rate is commonly referred as the _hazard rate_.

The Cox model is expressed by the hazard function denoted by $h(t)$. The hazard function can be interpreted as the risk of dying at time $t$.

It can be estimated as follow:

$$
h(t) = h_0(t) \times \exp (\beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_p x_p)
$$
where

* $h_0$ is the _baseline hazard_ corresponding to the value of the hazard if all of the $x_i$ are equal to zero.
* Each $\exp(\beta_i)$ are the _hazard ratios_. Id HR < 1the there is a reduction in the hazard and if HR > 1 then there is an increase in the hazard.


## Example in R

We will use the standard package `survival` for this and the package `survminer` which has some nice plotting functions.


```{r warning=FALSE, message=FALSE}
library("survival")
library("survminer")
```

Load the data

```{r}
data("lung")
head(lung)
```

and compute a univariate Cox model

```{r}
res.cox <- coxph(Surv(time, status) ~ sex, data = lung)
summary(res.cox)
```

For a multivariate Cox model it is a simple extension

```{r}
res.cox <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data =  lung)
summary(res.cox)
```

We can visualise the output as follows.

```{r}
fit <- survfit(Surv(time, status) ~ sex, data = lung)

# base R
plot(fit)

# ggplot2
ggsurvplot(fit, conf.int = TRUE,
           ggtheme = theme_minimal(), data = lung)
```

